04-elaboration exploration analyses

### 2021-05-21 Analysis/03-Elaboration_and_exploration.Rmd
Reorganised analyses

---
title: "Post MCMCglmm analyses"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

>>>>

---
title: title: "Bird beak's elaboration and exploration"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

# Measuring elaboration and exploration

## Per clade

## Per species

### 2021-05-21 Analysis/Older_ideas/01_data_preparation.Rmd
Reorganised analyses

>>>>

## Space visualisation

The full birdy space:

```{r, fig.height = 12, fig.width = 12}
## Loading the full space
shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")$PCscores
## Groups
group <- list("Psittaciformes" = trees_list$psittaciformes[[1]]$tip.label,
              "Charadriiformes" = trees_list$charadriiformes[[1]]$tip.label,
              "Others" = rownames(shapespace)[!(rownames(shapespace) %in% c(trees_list$psittaciformes[[1]]$tip.label, trees_list$charadriiformes[[1]]$tip.label))])
## Full space plot
plot(custom.subsets(shapespace, group = group), col = c("blue", "orange", "grey"), pch = c(19, 19, 21), main = "Bird beak shape space")
```

### 2021-05-21 Analysis/Older_ideas/02-elaboration_exploration_basics.Rmd
Reorganised analyses

```{r}
## Getting the base vectors for both groups
psita_vector <- apply(spaces_list$psittaciformes, 2, quantile, prob = c(0.25, 0.75))
chara_vector <- apply(spaces_list$charadriiformes, 2, quantile, prob = c(0.25, 0.75))

@@

```{r, print = FALSE}
## Making a top table
psita_tops_proj <- identify.tops(score = abs(psita_proj), data = spaces_list$psittaciformes)
psita_tops_reje <- identify.tops(score = psita_reje, data = spaces_list$psittaciformes)
chara_tops_proj <- identify.tops(score = abs(chara_proj), data = spaces_list$charadriiformes)
chara_tops_reje <- identify.tops(score = chara_reje, data = spaces_list$charadriiformes)

@@

>>>>

```{r}
## Getting the base vectors for both groups
psita_vector <- apply(spaces_list$psittaciformes, 2, quantile, prob = c(0.025, 0.975))
chara_vector <- apply(spaces_list$charadriiformes, 2, quantile, prob = c(0.025, 0.975))

@@

```{r, print = TRUE}
## Making a top table
psita_tops_proj <- identify.tops(score = abs(psita_proj), data = spaces_list$psittaciformes)
psita_tops_reje <- identify.tops(score = psita_reje, data = spaces_list$psittaciformes)
chara_tops_proj <- identify.tops(score = abs(chara_proj), data = spaces_list$charadriiformes)
chara_tops_reje <- identify.tops(score = chara_reje, data = spaces_list$charadriiformes)

@@

## Annex: some potentially interesting ratio thingy

```{r, fig.height = 6, fig.width = 6, eval = TRUE, print = TRUE}
## Loading the whole shape space
shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")$PCscores
## Getting the vector from that space
space_vector <- apply(shapespace, 2, quantile, prob = c(0.025, 0.975))
psita_space_proj <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ])
psita_space_reje <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
chara_space_proj <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ])
chara_space_reje <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
## Groups
group <- list("Psittaciformes" = trees_list$psittaciformes[[1]]$tip.label,
              "Charadriiformes" = trees_list$charadriiformes[[1]]$tip.label,
              "Others" = rownames(shapespace)[!(rownames(shapespace) %in% c(trees_list$psittaciformes[[1]]$tip.label, trees_list$charadriiformes[[1]]$tip.label))])
## Full space plot
plot(custom.subsets(shapespace, group = group), col = c("blue", "orange", "grey"), pch = c(19, 19, 21), main = "Bird beak shape space")
arrows(x0 = space_vector[1,1], y0 = space_vector[1,2],
       x1 = space_vector[2,1], y1 = space_vector[2,2], col = "black", lwd = 4)
```

```{r, fig.height = 8, fig.width = 8, eval = TRUE, print = TRUE}
par(mfrow = c(2, 2))
## The elaboration/exploration profiles
plot(psita_space_proj, psita_space_reje, pch = 19, col = "blue",
    xlab = "Elaboration", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
plot(chara_space_proj, chara_space_reje, pch = 19, col = "orange",
    xlab = "Elaboration", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")

## The elaboration/exploration profiles
plot(abs(psita_space_proj), psita_space_reje, pch = 19, col = "blue",
    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
plot(abs(chara_space_proj), chara_space_reje, pch = 19, col = "orange",
    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")
```

Meh... But we can also do ratios: the one between the within group elaboration/exploration and the total elaboration/exploration.
This can help us find the big time explorers/elaborators? (here: the one in the top 99% CI)

```{r, fig.height = 8, fig.width = 8, eval = TRUE, print = TRUE}
## Get the ratio
reje_ratio <- psita_reje/psita_space_reje
proj_ratio <- abs(psita_proj)/abs(psita_space_proj)

## Find the 99% top explorers/elaborators
top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
top_explo_names <- rownames(spaces_list$psittaciformes)[top_explo]
top_elab_names <- rownames(spaces_list$psittaciformes)[top_elab]

plot(proj_ratio, reje_ratio, pch = 19, col = "blue",
    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Psittaciformes")
text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
```

```{r, fig.height = 8, fig.width = 8, eval = TRUE, print = TRUE}
## Get the ratio
reje_ratio <- chara_reje/chara_space_reje
proj_ratio <- abs(chara_proj)/abs(chara_space_proj)

## Find the 99% top explorers/elaborators
top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
top_explo_names <- rownames(spaces_list$charadriiformes)[top_explo]
top_elab_names <- rownames(spaces_list$charadriiformes)[top_elab]

plot(proj_ratio, reje_ratio, pch = 19, col = "orange",
    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Charadriiformes")
text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
```

Maybe it's picking up some weird birds...

### 2021-05-21 Analysis/Older_ideas/03-ancestral_states_estimations.Rmd
Reorganised analyses

>>>>

```{r}
## TEMPORARY analyses: one matrix, one ace and one tree:
tmp_simple_matrices <- list(
    "psita" = make.dispRity(data = psita_max_ace[[1]], tree = trees_list$psittaciformes[[1]]),
    "chara" = make.dispRity(data = chara_max_ace[[1]], tree = trees_list$charadriiformes[[1]]))
save(tmp_simple_matrices, file = "../Data/Processed/tmp_simple_matrices.rda")
```

### 2021-05-21 Analysis/Older_ideas/04-elaboration_exploration_analyses.Rmd
Reorganised analyses

@@

@@

<!-- ## Annex: some potentially interesting ratio thingy

```{r, fig.height = 6, fig.width = 6, eval = FALSE, print = FALSE}
## Loading the whole shape space
shapespace <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")$PCscores
## Getting the vector from that space
space_vector <- apply(shapespace, 2, quantile, prob = c(0.25, 0.75))
psita_space_proj <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ])
psita_space_reje <- projections(spaces_list$psittaciformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
chara_space_proj <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ])
chara_space_reje <- projections(spaces_list$charadriiformes, point1 = space_vector[1,], point2 = space_vector[2, ], measure = "distance")
## Groups
group <- list("Psittaciformes" = trees_list$psittaciformes[[1]]$tip.label,
              "Charadriiformes" = trees_list$charadriiformes[[1]]$tip.label,
              "Others" = rownames(shapespace)[!(rownames(shapespace) %in% c(trees_list$psittaciformes[[1]]$tip.label, trees_list$charadriiformes[[1]]$tip.label))])
## Full space plot
plot(custom.subsets(shapespace, group = group), col = c("blue", "orange", "grey"), pch = c(19, 19, 21), main = "Bird beak shape space")
arrows(x0 = space_vector[1,1], y0 = space_vector[1,2],
       x1 = space_vector[2,1], y1 = space_vector[2,2], col = "black", lwd = 4)
```

```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
par(mfrow = c(2, 2))
## The elaboration/exploration profiles
plot(psita_space_proj, psita_space_reje, pch = 19, col = "blue",
    xlab = "Elaboration", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
plot(chara_space_proj, chara_space_reje, pch = 19, col = "orange",
    xlab = "Elaboration", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")

## The elaboration/exploration profiles
plot(abs(psita_space_proj), psita_space_reje, pch = 19, col = "blue",
    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Psitta explo/elaboration profile (total)")
plot(abs(chara_space_proj), chara_space_reje, pch = 19, col = "orange",
    xlab = "Elaboration (absolute)", ylab = "Exploration", main = "Chara explo/elaboration profile (total)")
```

Meh... But we can also do ratios: the one between the within group elaboration/exploration and the total elaboration/exploration.
This can help us find the big time explorers/elaborators? (here: the one in the top 99% CI)

```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
## Get the ratio
reje_ratio <- psita_reje/psita_space_reje
proj_ratio <- abs(psita_proj)/abs(psita_space_proj)

## Find the 99% top explorers/elaborators
top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
top_explo_names <- rownames(spaces_list$psittaciformes)[top_explo]
top_elab_names <- rownames(spaces_list$psittaciformes)[top_elab]

plot(proj_ratio, reje_ratio, pch = 19, col = "blue",
    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Psittaciformes")
text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
```

```{r, fig.height = 8, fig.width = 8, eval = FALSE, print = FALSE}
## Get the ratio
reje_ratio <- chara_reje/chara_space_reje
proj_ratio <- abs(chara_proj)/abs(chara_space_proj)

## Find the 99% top explorers/elaborators
top_explo <- which(reje_ratio >= quantile(reje_ratio, prob = 0.99))
top_elab <- which(proj_ratio >= quantile(proj_ratio, prob = 0.99))
top_explo_names <- rownames(spaces_list$charadriiformes)[top_explo]
top_elab_names <- rownames(spaces_list$charadriiformes)[top_elab]

plot(proj_ratio, reje_ratio, pch = 19, col = "orange",
    xlab = "Group/space elaboration", ylab = "Group/space exploration", main = "Charadriiformes")
text(cbind(proj_ratio, reje_ratio)[top_explo, ], labels = top_explo_names, pos = 4)
text(cbind(proj_ratio, reje_ratio)[top_elab, ], labels = top_elab_names, pos = 1)
```

Maybe it's picking up some weird birds...
 -->

>>>>

```{r, fig.height = 6, fig.width = 12}
## Centroid (tips) ancestor trajectory 
psita_results <- project.reject(tmp_simple_matrices$psita,
                                 vector = c("root", colMeans))
chara_results <- project.reject(tmp_simple_matrices$chara,
                                 vector = c("tips", "ancestor"))

@@

Measure the exploration/elaboration for all clades in the whole bird trees and see which ones are particularly low/high?

@@

## Ideas for getting the axis of "elaboration"

```{r, fig.height = 8, fig.width = 8}
library(dads)

## Simulate a pure-bd tree with a BM trait
data <- dads(stop.rule = list(max.living = 300), bd.params = list(speciation = 2), traits = make.traits(n = 3))
plot(data, main = "One trait through time")

## The node ages
ages <- tree.age(data$tree)
node_ages <- ages[which(ages$elements %in% data$tree$node.label),]

## A model of trait through time [TODO: use some time regression or something line that]
# time_trait <- cbind(time = rev(node_ages$ages), trait = data$data[data$tree$node.label, ])
# model <- lm(data$data[data$tree$node.label,] ~ 0 + node_ages$ages)
arrows(y0 = 0, x0 = max(node_ages$ages), y1 = mean(data$data[data$tree$tip.label, ]), x1 = 0, lwd = 3)
```

```{r}
library(rgl)
open3d()
plot(data, trait = c(1:2), use.3D = TRUE)
time_trait_arrow <- matrix(NA, 2, 3)
time_trait_arrow[1,] <- c(max(node_ages$ages),
                          0,
                          0)
time_trait_arrow[2,] <- c(0,
                          mean(data$data[data$tree$tip.label, 1]),
                          mean(data$data[data$tree$tip.label, 2]))
segments3d(time_trait_arrow, lwd = 3)

## 3D version
open3d()
plot(data, trait = c(1,2,3), use.3D = TRUE)
data_trait_arrow <- matrix(NA, 2, 3)
data_trait_arrow[1,] <- c(0,
                          0,
                          0)
data_trait_arrow[2,] <- c(mean(data$data[data$tree$tip.label, 1]),
                          mean(data$data[data$tree$tip.label, 2]),
                          mean(data$data[data$tree$tip.label, 3]))
segments3d(data_trait_arrow, lwd = 3)
